---
title: "Emerging Markets"
subtitle: "An exploration into four emerging markets. This paper will explore the correlation of FTSE indicies and GDP."
author: "Even Oscar Harlert"
date: 10/18/2023
format: pdf
---

## Introduction

Emerging markets can be great investment opportunity. In order to make good investment decisions and understand the risk and return for any investment you have to analyze historical data.

To measure countries development there are several metrics that can be useful. In this paper we will use GDP, **INSERT 3 MORE METRICS**. --\>

In this paper we will explore the correlation between stock market development and different development metrics in order to answer the following questions:

-   How correlated are the development metrics and the stock market indicies?

-   With the assumption that they are related, what are the reasons for divergences in the results?

-   

In order to answer the questions above this paper will analyze the data by deriving statistical metrics from the original data sets and test correlation between them .

## Data sets

There are two data sets acquired for this paper. The first one is downloaded from [refinitive/datastream](https://www.refinitiv.com/en). For more information regarding you can click this [link](https://www.refinitiv.com/content/dam/marketing/en_us/documents/fact-sheets/datastream-economic-data-macro-research-fact-sheet.pdf). This will be refereed to as dataset one. The data in the file are time-series spanning from 19. September 2003 to 19. September 2023. The measurements are done with a monthly interval. All data points are represented in US dollar.

The second data set is xxxxxxxxx. \<\-- get back to this

## Project librarys

This project are using a couple of librarys, they are the following:

```{r}
source("r_scripts/01libraries.R")
```

To make the code cleaner I've made a script folder and separated the required libraries in that file.

**\*\*INSERT TABLE OF THE FINAL VERSION OF LIBRARIES BEFORE EXPORTING THE REPORT \*\***

NB! We will add any library we add and check it at the end to assure that the project is not missing any dependencies.

## Data wrangling

We start our project out by downloading data from refinitive. Since the data is actually only accessable from the terminals located on floor five in the tower building at NMBU you would have to access these in order to downlaod the exact same data. I've provided a link to my set here (insert link to github). The data is stored as a excel file that has been stored in the project folder. To make it possible to work with in r we start by importing the file.

```{r}
index_emerging_markets <- read_excel("datasets/FTSE 20.09.23 - EOH.xlsx", skip = 6)

```

The data in this file is currently in absolute values. In order to compare them to each other we need to normalize them

The data is downloaded in US dollars and the 20 year frame is from 19 september 2003 and 19. September 2023.

From this point we want to normalize the data so we have a fair comparison point with the base of 100. In order to do so we are using the following line:

```{r}
normalized <- index_emerging_markets |>
  mutate(
    FTSEWorldN = FTSEWorld / (146.2 / 100),
    IndiaN = India / (332.5 / 100),
    EgyptN = Egypt / (59.04 / 100),
    ChinaN = China / (672.28 / 100),
    BrazilN = Brazil / (150.65 / 100),
    TaiwanN = Taiwan / (182.21 / 100),
  )
```

The code above adds 6 new columns where the data is normalized and starts at 100 on the 19. September 2003.

```{r}
final_desc_data_log |> head()
```

Here we now have a table of the number of the following:

-   n = Number of observations

-   mean = The average change

-   sd = standard deviation, the average deviation from the mean

-   skew = The tilt of the distribution curve

-   kurtosis = Messurment of the tail of the distrubtion curve.

This data will allow for us to compare the different data sets.

Future:

-   Check the Cov() and Cor()

Sidenote
